---
title: "Field Goal Probability Model"
subtitle: Kicker Rankings & Game Scenario Recommendations
output:
  html_document: 
    theme: sandstone
    fig_width: 5
    fig_height: 4
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
library(gt)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(nflfastR)

source('04_FG_Above_Expected.R')

# Load in play by play data for all games since the 2010 season
df <- load_pbp(2010:2024)

# Select columns relevant to field goals
df <- df %>% 
  select(
  # General Game Information 
  season, season_type, week, game_date, game_id, away_team, home_team, spread_line, total_line, 
  # Environmental Features
  time_of_day, location, stadium, weather, roof, surface, temp, wind, 
  # Play-Level Information
  play_id, posteam, defteam, posteam_type, game_half, qtr, play_clock, quarter_seconds_remaining, yards_gained,
  quarter_end, time, drive, down, side_of_field, yardline_100, yrdln, ydstogo, play_type, desc, series_result,
  # Situational Features
  score_differential, posteam_score, defteam_score, score_differential_post, ep, epa, wp, wpa, 
  home_timeouts_remaining, away_timeouts_remaining, timeout, timeout_team, 
  # Field Goal Information
  field_goal_attempt, field_goal_result, kick_distance, kicker_player_name, kicker_player_id
  )
```

------------------------------------------------------------------------

# Kicker Rankings

Below is a ranking of all kickers who attempted at least 10 field goals
during the 2024 season, from best to worst. These rankings evaluate
kickers using **Field Goals Above Expected (FGAx)**.

This metric goes beyond raw totals or accuracy - FGAx accounts for kick
difficulty by incorporating factors such as distance, weather, pressure
situations, and kicker ability. It provides a context-aware estimate of
**how many field goals a kicker made compared to how many they were
statistically expected to make**. This allows us to identify kickers who
are genuinely outperforming expectations, rather than simply benefiting
from a high volume of short-range attempts.

```{r}
# Create kicker rankings table
rankings <- fgax %>%
  ungroup() %>%
  filter(FG_attempted >= 10) %>%
  mutate(
    rank = row_number(),
    field_goal_percentage = round(FG_pct * 100, 2),
    expected_field_goals = round(xFG, 1),
    field_goals_above_expected = round(FGAx, 1),
    field_goal_percentage_above_expected = round(FGpct_Ax * 100, 2)
  ) %>%
  select(
    Rank = rank,
    Kicker = kicker_player_name,
    Team = posteam,
    `FGM` = FG_made,
    `FGA` = FG_attempted,
    `FG%` = field_goal_percentage,
    `xFG` = expected_field_goals,
    `FGAx` = field_goals_above_expected,
    `FG%Ax` = field_goal_percentage_above_expected 
  ) %>%
  gt() %>%
  tab_header(
    title = "Kicker Rankings: Field Goals Above Expected"
  ) %>%
  fmt_number(columns = c(`FG%`, `FG%Ax`), decimals = 1
  )

rankings
```

### Summary Insights

-   **Chris Boswell (PIT)** leads the league with 4.4 FGAx on 43
    attempts - a strong mark of elite performance across a sizable and
    varied workload.
-   **Brandon Aubrey (DAL)** also stands out with 3.8 FGAx, and made
    several successful kicks from more than 60 yards.
-   **Jake Bates (DET)** has been impressive in limited action, showing
    early promise as a rookie.
-   **Cameron Dicker (LAC)**, your kicker, ranks 7th overall with 2.2
    FGAx on 44 attempts. His ability to consistently exceed expectations
    at high volume is a clear positive and supports continued confidence
    in high-impact situations.

------------------------------------------------------------------------

------------------------------------------------------------------------

------------------------------------------------------------------------

# Game Scenario Recommendations

### Sunday Night Football in Week 15 \@ Kansas City

**Scenario Notes & Assumptions**: Week 15 \@ Kansas City (Sunday Night
Football)

-   Mid-December, road game in Kansas City

-   Clear conditions with 20 mph wind gusts (no precipitation).

-   Excluding final 2 minute scenarios game-tying or game-winning
    attempts.

-   Cameron Dicker or another average, non-rookie NFL kicker is making
    the attempt

-   Go-For-It Assumption: 4th-and-2 situation conversion gains a new set
    of downs, not a touchdown (not 4th-and-goal).

-   Turnover on Downs Assumption: the opponent gains possession in their
    own territory, with expected points from that field position set to
    0 once past midfield.

    ------------------------------------------------------------------------

```{r, warning = FALSE, message = FALSE, eval=FALSE}
# Predict field goal probability at a range of distances 
kick_distances <- seq(21, 70, by = 1)

# Create data frame for predictions
scenario_df <- data.frame(
  kick_distance = kick_distances,
  wind = rep(20, length(kick_distances)),
  precipitation = rep(0, length(kick_distances)),
  season = rep(2024, length(kick_distances)),
  last_two_minutes = rep(0, length(kick_distances)),
  kick_to_win = rep(0, length(kick_distances)),
  adj_fg_pct = rep(
    all_data %>% 
      ungroup() %>%
      filter(season == 2024, week == 15) %>%
      summarise(avg_fg_pct = mean(adj_fg_pct, na.rm = TRUE)) %>%
      pull(avg_fg_pct),
    length(kick_distances)),
  is_rookie = rep(0, length(kick_distances))
) 

# Predict FG probability at each distance 
scenario_df$fg_probability <- predict(football_model, newdata = scenario_df, type = "response")

# Convert to expected points at each distance 
scenario_df$exp_points <- (scenario_df$fg_probability * 3) + ((1 - scenario_df$fg_probability) * 0)

# Add a small penalty to longer kicks, in order to lean a little conservative 
scenario_df <- scenario_df %>%
  mutate(penalty = (kick_distance - 40) * 0.005, 
         fg_probability_conservative = pmax(fg_probability - penalty, 0),
         exp_points = fg_probability_conservative * 3)

# Define a range of yardlines
yardlines <- seq(1, 60, by=1)

# Compute expected points using nflfastR function
ep_df <- data.frame(
  season = rep(2024, length(yardlines)),
  home_team = rep("KC", length(yardlines)),
  posteam = rep("LAC", length(yardlines)),
  roof = rep("outdoors", length(yardlines)),
  half_seconds_remaining = rep(1800, length(yardlines)),
  yardline_100 = yardlines - 2,
  down = rep(1, length(yardlines)),
  ydstogo = rep(10, length(yardlines)),
  posteam_timeouts_remaining = rep(3, length(yardlines)),
  defteam_timeouts_remaining = rep(3, length(yardlines))
)

# Compute expected points from going for it at each yardline
ep <- calculate_expected_points(ep_df) %>%
  mutate(yardline_100 = yardline_100 + 2,
         ep = ep/2) %>%
  select(distance = yardline_100, go_expected_points = ep)

# Identify the point where ep from going for it > ep from kicking a fg 
scenarios <- scenario_df %>%
  mutate(
    yardline_100 = kick_distance - 18,
    yard_line = 100 - yardline_100
  ) %>%
  select(
    distance = yardline_100,
    yard_line,
    fg_expected_points = exp_points
  ) %>%
  left_join(ep, by = "distance") %>%
  mutate(
    fg_expected_points = round(fg_expected_points, 1),
    go_expected_points = round(go_expected_points, 1)
  )

min_ydl <- scenarios %>% 
  filter(fg_expected_points < go_expected_points) %>% 
  summarise(min_yardline = min(distance))

scenario_1 <- scenarios %>%
  pivot_longer(
    cols = c(fg_expected_points, go_expected_points),
    names_to = "type",
    values_to = "expected_points"
  ) %>%
  mutate(type = dplyr::recode(type, 
                       fg_expected_points = "Field Goal Attempt",
                       go_expected_points = "Go For It"))

# Plot expected points for each decision and yardline
yardline_plot <- ggplot(scenario_1, aes(x = distance, y = expected_points, color = type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_vline(xintercept = as.numeric(min_ydl), color = '#001433', linetype = "dotted", linewidth = 0.7) +
  scale_color_manual(values = c("Field Goal Attempt" = "#0080C6", "Go For It" = "#F0AE00")) +
  labs(
    title = "Expected Points by Distance",
    x = "Kick Distance/Yardline (yards)",
    y = "Expected Points",
    color = "Decision"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    legend.position = "top"
  ) + scale_x_continuous(breaks = seq(0, 60, by = 5))

yardline_plot

```

#### **Recommendation**

> In Week 15 at Kansas City on Sunday Night Football --- with 20 mph
> wind gusts expected --- we recommend attempting a field goal when the
> line of scrimmage is at or inside the opponent's 36-yard line. This
> corresponds to a **54-yard field goal attempt or shorter**.
>
> This is the point at which, given the conditions, the expected points
> from a field goal attempt exceed the expected value of going for it on
> 4th-and-2, assuming a 50% conversion rate. Beyond this distance, the
> chance of making the kick drops enough that the expected return no
> longer justifies the attempt compared to the risk-adjusted value of
> going for it.

------------------------------------------------------------------------

------------------------------------------------------------------------

### New, Free Agent Kicker in Week 7 \@ Home

**Assumptions**

-   The free agent, off-the street-kicker signed on Tuesday is a rookie,
    is equivalent to the 33rd best kicker, worse than all kickers
    currently on rosters, but not a high-schooler.

```{r}
# Predict field goal probability at a range of distances 
kick_distances <- seq(11, 60, by = 1)

# Create data frame for predictions
scenario_2 <- data.frame(
  kick_distance = kick_distances,
  wind = rep(0, length(kick_distances)),
  precipitation = rep(0, length(kick_distances)),
  season = rep(2024, length(kick_distances)),
  last_two_minutes = rep(0, length(kick_distances)),
  adj_fg_pct = rep(all_data %>% 
                     filter(week == 7 & season == 2024) %>% 
                     summarise(min = min(adj_fg_pct)) %>% 
                     pull(min), 
                   length(kick_distances)),
  is_rookie = rep(1, length(kick_distances))
) 

# Predict FG probability at each distance 
scenario_2$fg_probability <- predict(football_model, newdata = scenario_2, type = "response")

# Convert to binary predictions, using 0.9 threshold to maximize precision/recall and confidence
scenario_2$fg_prediction <- if_else(scenario_2$fg_probability > 0.75, 1, 0)

# Maximum distance to attempt
max_dist <- scenario_2 %>% filter(fg_prediction == 1) %>% summarise(max(kick_distance))
```

------------------------------------------------------------------------

#### **Field Goal Target Line Recommendation** 

> Given that we're working with a free agent kicker signed earlier this
> week, and factoring in the uncertainty that comes with limited
> practice and no recent NFL game reps, our recommendation is:
>
> **Attempt field goals of 46 yards or less**.
>
> From this range, we're 75% confident the kicker will convert. Beyond
> that, confidence drops off quickly, and attempting a field goal is not
> recommended unless the situation demands it.
